---
title: "Cheat Sheet"
date: "`r Sys.Date()`"
author: "Justin Meisenhelter"
output: 
  rmdformats::downcute:
    highlight: tango
    df_print: paged
    collapsed: false
---

# Python  

## Indexing   

### Indexing a String  

`s[start_position:end_position:delta]`
```{python}
word = 'abcdefghijk'
word[-1:-3:-1]
```


### Reversing a string  

```{python}
word[::-1]

```
### String Methods  

https://www.w3schools.com/python/python_ref_string.asp
```{r, eval=FALSE}
word.upper()
word.lower()
word.strip()
word.replace()
word.split()
```


## Printing   

### Modulous Operator  

```{python}
print( "hello, %f!" %2017)
print( "hello, %i!" %2017)
print( "hello, %.2f!" %2017)
print( "hello, %s!" %'2017')

```
## Functions

```{python}
def square(x):  
    return x**2
```

### Default Arguments
```{python}
def polynomial(x, a, b=0, c=0, d=0):  #assigning default values
    result = a + b * x**1 + c * x**2 + d * x**3
    return result

```
## Control Flow  

### Else If  

```{python}
a=6
if type(a)==str:
    print('it is a string')
else:
    if a%2==0:
        print('it is an even number')
    else:
        print('it is an odd number')
```
```{python}
if type(a)==str:
    print('it is a string')
elif a%2==0:
    print('it is an even number')
else:
    print('it is an odd number')
```
### While Loops  

```{python}
x = 1
while x <= 10:
    if x % 2 == 0 and x % 3 != 0:
        print (x)
    x = x + 1

```
### for loops  

```{python}
for x in range(11):
    if x % 2 == 0 and x % 3 != 0:
        print (x)
```
## Handling Errors/Exceptions  

Python's **try** and **except** can provide ways to handle exceptions.

Exceptions raised by statements in the body of try are handled by the except statement and execution continues with the body of the except statement.  

```{python}
def divide(x, y):
    try:
        result =  x / y
    except ZeroDivisionError:
        result = 'INFINITY'
    except TypeError:
        result = divide(float(x), float(y))
    return result
```
## Lists

### Defining Lists
```{python}
x=[1,2,3,4,'1','a','hello']
print(x)
```
```{python}
#create an empty list
y = list()
```

### List Methods
```{python}
x.index('hello')
```
```{python}
x.append(55)
```




### List comprehension
```{python}
x=[200,400,23,50,1,23,54,66,88,97,20]
y=[i for i in x if i%2==0]
y
```
Write a list comprehension that takes a list of strings and returns the capitalized version of those strings if they are greater than 2 and fewer than 12 characters in length.
```{python}
names=['Barry White','aj','tom','John Foster Wallace',
       'Billy Bud','Yifan Ghost Song']
[i.upper() if len(i) > 2 and len(i) <12 else i for i in names]
```

## Tuple  

Tuples are similar to lists in many ways, except they can't be modified once you've created them.  

Tuples are defined between two parentheses instead of brackets.  

Tuples are iterable but immutable.  
```{python}
tup=("hi", True, 1.0, 2)
tup

```
## Set  

A set is an unordered sequence of unique values. 
### Creating an empty set
```{python}
z=set() #creating an empty set
z
```
### Set Function
```{python}
s={1,2,3,3,3,4}
t={8,4,6,2}
s.difference(t)
s-t
s.intersection(t)
s&t
s.union(t)
s|t
s^t
# all non shared items in set
```
## Dictionary  

A dictionary is a set of keys with associated values. The key must be hashable, but the value can be any object. It is also known as a hash map, hash table or set of key-value pairs.  
```{python}
thing={}
thing['key1']='val1'
thing[2]='val2'
thing[3]=[3,2,1]

```
### Dictionary functions
```{python}
thing.keys()
thing.values()
thing.items()
```
### Word Frequency Histogram
```{python}
words=['hi','hello','hi','hi','yo','whazzup','hey','hey']
hist={}
for i in words:
    if i in hist:
        hist[i] += 1
    else:
        hist[i] = 1
hist

```
## Arrays  

```{python}
import numpy as np
nested_lst1 = [[-1,2,1,2],[5,-6,5,6],[7,8,-7,8]]
nested_lst2 = [[1,2,1,2],[5,6,5,6],[7,8,7,8]]
multi_ary1 = np.array(nested_lst1)
multi_ary2 = np.array(nested_lst2)
higher_dim = np.array([multi_ary1, multi_ary2])
higher_dim
```
### Array Functions
```{python}
np.arange(10)
np.linspace(0,10, 51) #(start, end, no of elements)
np.ones((3,2)) #(array of ones of dimension 3x2)
np.zeros((4,3,2)) #array of zeroes of dimension 4x3x2)
high_x1 = np.array([[1,2,3],[4,5,6]])
high_x1.shape
x = np.arange(8)
x.reshape([2,4], order='F') #reshapes array to 2x4 fills by columns
```
## Matrix
```{python}
x = np.matrix([3,2])
print(type(x))
print(x.shape)
x
```
### Matrix Multiplication  

The type matrix is for linear algebra, in which the matrix multiplication is always **“a row times a column”**:

$$
(1,2,3) \times \left(
\begin{array}{c}
1\\
2\\
3\\
\end{array}
\right)
= (1+4+9) = 14
$$
### Identity Matrix  
```{python}
Id = np.matrix(np.eye(3))
Id
```
### Inverse Matrix

The identity matrix is analogous to the number one.  One is referred to as the multiplicative identity, since any number multiplied by one remains unchanged. A nonzero number has a multiplicative inverse, or reciprocal.  The product of a number and its multiplicative inverse is the multiplicative identity, one. 

**For real scalars:**

- nonzero number: n
- multiplicative identity: 1
- multiplicative inverse: $\frac{1}{n}$
$$n \times 1 = n$$

$$n \times \frac{1}{n}=1$$

The analogy for a square matrix is called an inverse matrix.  A nonsingular square matrix has a multiplicative inverse, or inverse matrix.  The product of a matrix and its inverse matrix (in either order) is the identity matrix. 

**For real-valued square matrices:**

- nonsingular matrix: $M$
- identity matrix: $\textbf{1}$
- inverse: $M^{-1}$
$$M \cdot \textbf{1} = M$$

$$M \cdot M^{-1}=\textbf{1}$$

$$M^{-1} \cdot M=\textbf{1}$$  

- How do you find a tuple of `x`, `y` and `z` satisfying all the equations below?
$$
\begin{eqnarray}
3x +2y -z &= 1\\
2x -2y +4z &=-2\\
-x +\frac{1}{2}y- z&= 0
\end{eqnarray}
$$

Observe that the equation can be written as:

$$
\begin{pmatrix}
3 &2  &-1 \\ 
2 &-2  &4 \\
-1&\frac{1}{2}&-1
\end{pmatrix}
\times
\begin{pmatrix}
x\\y\\z
\end{pmatrix}
=
\begin{pmatrix}
1\\-2\\0
\end{pmatrix}
$$

$$\begin{pmatrix}
x\\y\\z
\end{pmatrix}=
\begin{pmatrix}
3 &2  &-1 \\ 
2 &-2  &4 \\
-1&\frac{1}{2}&-1
\end{pmatrix}^{-1} \times
\begin{pmatrix}
1\\-2\\0
\end{pmatrix}$$  


```{python}
M = np.matrix([[3, 2, -1],[2, -2, 4],[-1, 0.5, -1]])
M2 = np.matrix([[1], [-2], [0]])
M.I * M2
```



## Classes  

* `Person` is the **class**
* `name` and `age` are each an **object attribute**
* `say` is an **object method**
```{python}
class Person():
    '''This is the documentation of the class person ... here'''
    
    def __init__(self,a_name,a_age):#constructor
        self.name=a_name
        self.age=a_age
        
    def intro(self):
        print("My name is",self.name,"and I am",self.age,"years old")

```
## Regular Expressions
https://learnbyexample.github.io/python-regex-cheatsheet/


## Files  

### Opening Files
```{python, eval=FALSE}
f = open('fooz.txt', 'r') 
f.read()
f.close()
# reads entire file as a string
```
### Reading Files
```{python, eval = FALSE}
f = open('fooz.txt', 'r')
lines=f.readlines()
print(lines)
f.close()
# reads each line as list
```
### Iterating Files  
```{python, eval = FALSE}
f = open('fooz.txt', 'r')
for i in f:
    print(i)
f.close()
```
### Writing files
```{python, eval = FALSE}
f.write('this is some text.\nThis is another line.\n')
f.write('oh snap, more lines.\nIt\'s a line.\n')
f.close()
```
## Random Sampling  

The `numpy.random` submodule provides random-variable generator. Some are listed below:

- `randn(d0, d1, ..., dn)`: Each entry is from the “standard normal” distribution.

- `rand(d0, d1, ..., dn)`: Random array with a given shape where each element is from Uniform [0, 1].

- `randint(low, high, size)`: Random integers ranging from low (inclusive) to high (exclusive).

- `random_integers(low, high, size)`: Random integers between low and high, both inclusive.

- `random_sample(size)`: Random floats in the half-open interval [0.0, 1.0).

- `choice(a[, size, replace, p])`: Generates a random sample from a given 1-D array. By default, `replace=True`.  
```{python}
import numpy as np
np.random.randn(10)
```
## Scipy
```{python, eval=FALSE}
import numpy as np
from scipy import stats

from sklearn import datasets
iris_raw = datasets.load_iris()
```
```{python, eval = FALSE}
iris = iris_raw.data
iris[0:8,:]
sepal_len = iris[:,0] # gets first column, sepal length
sepal_wid = iris[:,1] # gets 2nd column, sepal width
petal_len = iris[:,2] # gets column 3, petal length

```

## Statistical Functions  
```{python, eval=FALSE}
stats.describe(sepal_len)
np.unique(species, return_counts = True) 
stats.ttest_1samp(petal_len, 10) # 1 sample t test
stats.ttest_ind(sepal_len, petal_len) #2 sample t test
stats.f_oneway(sepal_len, sepal_wid, petal_len) # ANOVA

```
## Distributions

```{python, eval=FALSE}
my_binom = stats.binom(6, 0.5) # binomial distribution
my_binom.rvs(10) # random sampling
my_binom.pmf(3) # returns likelyhood
my_norm = stats.norm(3, 1) # normal distribution with mean = 3, stdev = 1
my_norm.cdf(4) #culmulative density object
my_norm.pdf(4)) # probability density function

```
## PANDAS

### Series
```{python}
import pandas as pd
obj = pd.Series(['a', 'b', 'c', 'd'])
obj

```
#### Slicing
```{python}
obj[0]
obj.index
obj.values

```
#### Convert Dictionary to series
```{python}
dict_ = {1: 'a', 2: 'b', 3: 'c', 4: 'd'}
obj3 = pd.Series(dict_)
obj3
obj3.to_dict() # and back to dictionary

```
### Dataframes

#### Creating a dataframe  
```{python}
# From Dictionary
data = {'commodity': ['Gold', 'Gold', 'Silver', 'Silver'],
        'year': [2013, 2014, 2014, 2015],
        'production_Moz': [107.6, 109.7, 868.3, 886.7]} 


df = pd.DataFrame(data)
df
# from nested list
df_2=pd.DataFrame([[107.6, 'Gold', 2013],
                   [109.7, 'Gold', 2014],
                   [868.3, 'Silver', 2014],
                   [886.7, 'Silver', 2015]], 
                    columns=['production_Moz','commodity','year'])
df_2
```

#### DF Values
```{python}
df.values
df.index
df=df.set_index('commodity') # sets index to commodity column
df=df.reset_index()
df.columns
df.index

```

#### DF slicing
```{python}
# how to get columns
df.commodity
df['commodity']
df.iloc[1]
df.loc[:,'commodity'] #row : Col
df[['commodity']] # returns a df
df['commodity'][0] # returns comodity column row 0
df[['commodity', 'year']]
df.loc[df['commodity'] == 'Gold', :]
df.loc[:, df.loc[0] == '2013']

```
#### DF Functions
```{python, eval = FALSE}
df.sort_values('value') # column to sort on
df.concat(df1, df2)
pd.merge() # df1, df2, how(left, inner), on(column to merge on)

```
#### Group
```{python, eval=FALSE}
df.groupby('colname')
# groups are iterable
for item in group:
  print item
# make a grouypby dictionary
group_dict = dict(iter(group))
group.agg(['count', 'mean', 'min', 'max', 'sum', 'std'])
```


#### Read from file
```{python, eval = FALSE}
df=pd.read_csv('foo.csv')
df

```
#### DF Null values
```{python, eval = FALSE}
df.isnull() # returns t/f dataframe
np.sum(df.isnull()) # sums nulls per column
one      2
two      2
three    2
four     1
dtype: int64
np.sum(df_miss.isnull(), axis=1) # sums nulls per row
df.dropna(axis=0, how='any') # drops rows with NA values
df.fillna(0) # fills NA values
df['one'].fillna(df_miss['one'].mean()) # fills all NAs in 'one' column with the mean
df.interpolate() # fills NA values with a linear regression of surrounding values
```
## MatplotLib
```{python, eval = FALSE}
import matplotlib.pyplot as plt
x=[5,5,5,10,10,30,30,20,25,40]
plt.hist(x)
plt.show()
values=[1628701 ,2582830, 1432132, 476179 , 2278906]
labels=['Manhattan','Brooklyn','Bronx','Staten Island', 'Queens']
plt.pie(values,labels=labels)
plt.boxplot(x)
x=[7,8,9,11]
y=[9,4,6,7]
plt.bar(x,y);
plt.barh(x,y);
plt.scatter(x,y);

```
### Plotting Functions
```{python, eval = FALSE}

X=np.linspace(0,4*3.14,101)
Y=np.sin(X)
plt.plot(X,Y,'.')
```
### Plotting Features
```{python, eval = FALSE}
plt.figure(figsize=(12,6)) #figsize must be first declaration
plt.plot(x,y,'g.-',markersize=20, linewidth=4,alpha=0.4,label='data1')
plt.plot(x,y2,'.--',color='red',markersize=10,alpha=0.7,linewidth=0.3,label='data2')
plt.legend(loc=1)
plt.title('Example Plot')
plt.xlabel('horizontal axis')
plt.ylabel('vertical axis')
plt.xlim(6,12)
plt.ylim(0,10)
plt.xticks([6,9,12])
plt.yticks([0,4,8])
plt.text(10,2,'some text',color='blue',size=14)
plt.show()
```

## Seaborn
```{python, eval = FALSE}
import seaborn as sns
titanic=sns.load_dataset('titanic')
```
```{python, eval = FALSE}

sns.countplot(x="deck", data=titanic);
plt.hist(titanic['deck'].dropna());
sns.stripplot(x="sex",y="age",data=titanic)
sns.swarmplot(x="sex",y="age",data=titanic)
sns.barplot(x="sex",y="survived", data=titanic)
sns.barplot(x="sex",y="survived", hue='class', data=titanic)
```



# R  

## Markdown  

### Formatting  
For basic formatting for your output HTML document    
(how to update your YAML Preamble):  

https://bookdown.org/yihui/rmarkdown/html-document.html  


### Themes:  

https://www.datadreaming.org/post/r-markdown-theme-gallery/  


### Code Chunk   

https://yihui.org/knitr/options/  


### Advanced Themes:  

https://cran.r-project.org/web/packages/rmdformats/readme/README.html 


### LaTex:  

https://joshua.smcvt.edu/undergradmath/undergradmath.pdf  


### HTML   

colors, fonts, etc... (most tags will not work if placed around codechunks, or will not work as expected)  

https://daringfireball.net/projects/markdown/syntax  

## Functions
```{r}
cost_per_person <- function(people, bill) {
  tax = bill * 0.07
  tip = bill * 0.18
  total = bill + tax + tip
  costPP = total / people
  return(costPP)
}
```
```{r}
cost_per_person(5, 221.78)
cost_per_person(4, 221.78)
```
### Default Arguments  
```{r}
xyz = function(x=1, y, z=1) {
  return(x * 100 + y * 10 + z)
}
```

### Built In Functions
```{r}
set.seed(0) # For reproducible results
x = rnorm(1000, mean=2) # Random sample from a normal distribution

mean(x) # mean of x
median(x) # median of x
var(x) # variance of x
sd(x) # standard deviation of x
quantile(x, c(0.25, 0.50, 0.75)) # quantiles of x
sum(x) # sum of x
```

## Working Directory
```{r}
getwd()
```
```{r}
#setwd("C:/Users/cmmaf/Desktop/RDA-50/1/Lecture")
#setwd("C:/Users/cmmaf/Desktop/RDA-50")
```

## Reading Files
```{r}
#full_path = "C:/Users/cmmaf/Desktop/RDA-50/1/Lecture/data/CPS1988.csv"
#CPS1988 = read.csv(full_path, header=TRUE)

#relative_path = "data/CPS1988.csv"
#CPS1988 = read.csv(relative_path, header=TRUE)
library(AER)
data("CPS1988")

dim(CPS1988)
```

## Data Frames

### Useful Methods
```{r}

head(CPS1988)
str(CPS1988, vec.len=5)
head(CPS1988$region)
```

### Slicing Data Frames
```{r}
str(CPS1988, vec.len=1)
```
```{r}
CPS1988$wage[1:8]
```
```{r}
CPS1988[["wage"]][1:8]
```
```{r}
CPS1988["wage"]
```

```{r}
CPS1988[c("wage", "education")]
```

```{r}
# Use $ to get individual columns out as vectors
CPS1988$wage[1:6]
CPS1988$education[1:6]
```

```{r}
# We can use brackets as well, however since df's have 2 dimensions df[ROWS, COLUMNS]

# Returns a df, first 10 rows and columns 2 and 4
CPS1988[1:10, c(2,4)]
```

```{r}
CPS1988[1:10, c('education', 'ethnicity')]
```





```{r}
# All rows but for the educ and exper columns
CPS1988[c('education', 'experience')]
#CPS1988[ , c('education', 'experience')]
```

```{r}
# If you get only one column, it will be a vector
class(CPS1988[, 1])
class(CPS1988[, 1:2])
```
```{r}
CPS1988["wage"]
```

```{r}
CPS1988[1]
```

## Lists
```{r}
myVec = 1:10
```
```{r}
myMat = matrix(1:12, ncol=3)
```

```{r}
myDf = CPS1988[1:10, ]
```

```{r}
myList = list(myVec, myMat, myDf)
myList
namedList = list(A=myVec, B=myMat, C=myDf)
namedList$D = pi

namedList['E'] = exp(1)

namedList
```

## Control Flow

### IF  
```{r}
num = 5
# Note: %% is the Modulo (remainder) operation
if (num %% 2 != 0) {
  cat(num, 'is odd')
}
```
```{r}
num = 5

if (num %% 2 != 0) {
  cat(num, 'is odd')
  # more commands
} else {
  cat(num, 'is even')
  # more commands
}
```
```{r}
set.seed(0)
age = sample(0:100, 20, replace=TRUE)
age
res = ifelse(age > 70, 
             'old', 
             ifelse(age <= 30, 
                    'young', 
                    'mid'))
res
```
### FOR
```{r, eval=FALSE}
library(readr)
sign_data = read.csv("TimesSquareSignage.csv", header=TRUE)
obs = nrow(sign_data)
for (i in 1:obs) {
  if (is.na(sign_data$Width[i])) {
    cat('WARNING: Missing width for sign no.', i, '\n')
  }
}
```
### While
```{r, eval = FALSE}
i = 1
while (i <= obs) {
  if (is.na(sign_data$Width[i])) {
    cat('WARNING: Missing width for sign no.', i, '\n')
  }
  i = i + 1
}
```
## Dplyr

### Tibble
```{r}
library(dplyr)
births_df = read.csv("C:/Users/jmeis/RDA_054/Lecture 2/data/births.csv", stringsAsFactors=FALSE)
bnames_df = read.csv("C:/Users/jmeis/RDA_054/Lecture 2/data/bnames.csv", stringsAsFactors=FALSE)
births = as_tibble(births_df)
bnames = as_tibble(bnames_df)
```
### Select, Filter, Arrange
```{r}
bnames %>%
  select( year, name, prop) %>%
  filter( name == 'Justin') %>%
  arrange(desc(prop)) %>%
  head(2)
```
### Mutate, Transmute, Summarise
```{r}
bnames %>%
  mutate(., newCol = year *2) %>%
  transmute(newCol = newCol/2) %>%
  summarise(sum = sum(newCol/n()))
```
### Joins
```{r}
x = data.frame(
  name = c("John", "Paul", "George", "Ringo", "Stuart", "Pete"),
  instrument = c("guitar","bass","guitar","drums","bass","drums"))
y = data.frame(
  name = c("John", "Paul", "George", "Ringo", "Brian"),
  band = c(TRUE, TRUE, TRUE, TRUE, FALSE))
left_join(x, y, by = "name")
right_join(x, y, by = "name")
inner_join(x, y, by = "name")
full_join(x, y, by="name")
semi_join(x, y, by = "name")
anti_join(x, y, by = "name")
```
### Group_by

```{r}
totals = bnames %>%
  group_by(name) %>%
  summarise(total = n())

head(totals)
```



## Shiny

### Links to resources: 
HTML Tags
https://shiny.rstudio.com/articles/tag-glossary.html
Layout Guide
https://shiny.rstudio.com/articles/layout-guide.html
Functions
https://shiny.rstudio.com/reference/shiny/latest/
Widgets
https://shiny.rstudio.com/gallery/widget-gallery.html

### Basic Layout (ui.r)
```{r, eval=FALSE}
fluidPage(
    titlePanel(...),
    sidebarLayout(
      sidebarPanel(...),
      mainPanel(...)
)
)
```
```{r, eval=FALSE}
library(shiny)
library(lubridate)

# Define UI for application that draws a histogram

fluidPage(
  titlePanel("NYC Flights 2014"),
  sidebarLayout(
    sidebarPanel(
     
      selectizeInput(inputId = "origin",
                    label = "Departure airport",
                     choices = unique(flights$origin)),
      selectizeInput(inputId = "dest",
                     label = "Arrival airport",
                     choices = unique(flights$dest)),
      selectizeInput(inputId = "month",
                     label = "Month",
                     choices = unique(flights$month)
    ),
    mainPanel(tabsetPanel(
      tabPanel('Number of Flights', plotOutput("count")),
      tabPanel('Average Delay', plotOutput("delay"))
    ))
  )
)
```


### Basic Server (server.r)
```{r, eval=FALSE}
function(input, output) {
}
```
```{r, eval=FALSE}
library(shiny)
library(dplyr)
library(ggplot2)
library(tidyverse)

function(input, output) {
  output$count <- renderPlot(
    flights %>%
      filter(origin == input$origin & dest == input$dest & month == input$month) %>%
      group_by(carrier) %>%
      count() %>%
      ggplot(aes(x = carrier, y = n)) +
      geom_col(fill = "lightblue") +
      ggtitle("Number of flights"))
  output$delay <-renderPlot(
    flights %>%
      filter(origin == input$origin & dest == input$dest & month == input$month) %>%
      group_by(carrier) %>%
      summarise(mean_arr = mean(arr_delay), mean_del = mean(dep_delay)) %>%
      gather(key = metric, value = delay, mean_arr, mean_del)  %>%
      ggplot(aes(x = carrier, y = delay, fill = metric )) +
      geom_col(position = "dodge2") +
      ggtitle('Average Delays')
      
  )
  
  
}
```
### Global .r
```{r, eval =FALSE}
#for above example
library(data.table)
flights <- fread(file = "./flights14.csv")

```
# HTML  

## Workflow
Start new document index.html with !DOCTYPE

## Parts of HTML Document  
\<title\> \<\/title\>
\<head\> \<\/head\>  
contains technical info about meta structure of doc  
\<body\> \<\/body\>  
content goes here 
Lorem# is number of lorem ipsum words to create

## Headers and Paragraphs and divs  
\<h1\> \<\/h1\>   
\<p\> \<\/p\>  
\<div\> \<\/div\> 

## Insert an image  
\<a href="about.html"\>About\<\/a\>

# CSS  
for external css create a style.css page  

## Inline CSS  
\<h1 style="color: olive;font-size: 3rem;"\>hello world\<\/h1\>

## Internal CSS
\<style\>
h1{  
  color:blue;  
  font-size: 3rem;  
}  
will change the apperance of all \<h1\> elments  
will will with spans, body, divs, paragraphs, etc...
\<\/style\>

## External CSS  
### Element selectors and grouping  
h1,h2{  
  color: red;  
}  
### IDs  
/#heading{  
  background:black;  
}  
\<h1 id="heading"\>lorem ipsum \<\/h1\>  

### Class Selectors 
.green{  
  color:green;  
}  
\<h3 class="green"\>I'm green\<\/h3\>  

### Universal Selector  
*{  
  background-color:ivory  
}  
Universal selector will be the top level css style, unless overwritten by subsequent styles  

### RGBA 0-1 Opacity/Transparency
\#\<tag\>{  
  background:rgba(0,0,0,.25)  
  \<(r value, g value, b value, opacity)\>  
}  

### Font size, width, height, pixels
{  
font-size: 30px;  
width: 200px;  
height: 300px;  
}  

### Percent Value Relative  
{  
width: 50%;  
height: 50%;  
}  

# SQL  

## comments  
\-\- Sytanx for single line comments  
/** multiline comment **/  

## _Commands_

### _SHOW \<DATABASES\> or SHOW \<database, or table\>;_
shows all databases on server, or shows DB or table

### _USE \<database\>_ 
will append database. to every table reference for example:  
`USE movies_db;`
is the same as   
`FROM movies_db.tables;`  

### _SELECT \<rows\>_
selects all rows  

### _FROM \<database\>_  
noun of the select querry  

### _WHERE ..._   
indicates the condition or conditions that rows must
satisfy to be selected.  
```
SELECT movie_title,
budget
FROM movies
WHERE budget > 1000000000;

```

### _GROUP BY ..._  

### _HAVING ..._  

### _ORDER BY ..._   

### _LIMIT ..._  
```
SELECT column_names  
FROM tabel_names  
LIMIT \<offset\> \<count\>;  
```
### _AS ..._  

### _CASE WHEN \<condition\> THEN \<result\> ELSE \<result\> END_ 
```
SELECT movie_title,   
CASE   
       WHEN title_year > 2014 THEN 'new' 
       WHEN title_year BETWEEN 2012 AND 2014 THEN 'new-ish'   
ELSE 'old'  
END AS movie_age  
FROM movies;  
```
### _DESCRIBE \<object\>_  

### _Column Alias_
Select <column name> AS <column alias name>

##  Mathmatical Operators
\+  
\-  
\*  
/  
Round()  

## Comparison Operator  
= , equal to  
\> , Greater than  
\< , less than  
\>= , greater than or equal to  
\<= , less than or equal to  
\<\> , \!= , not equal to  

## Logical Operators
### _AND_  
Evaluates two conditions and
returns True if both are true.  

### _OR_  

Evaluates two conditions and
returns True if at least one is
true.   

### _BETWEEN...AND... _  

Tests value against an inclusive
range of values. 
```
SELECT movie_title,  
title_year BETWEEN 2010 AND 2012 AS year_range  
FROM movies;  
```
### _IN_  
Tests value against a list of
values and returns True if value
matches at least one in the list. 
```
SELECT movie_title, 
title_year IN (2010, 2012, 2014) AS year_range,  
country IN ('USA', 'UK') AS country_range  
FROM movies;  
```

### _IS NULL_  
Tests value to see if it is null.
Return True if null.  

### _NOT_   
Reverts False to True, and True
to False.    

## String Manipulation  

### _Wildcard Comparisons_  
_LIKE_ operator in combo with wildcards are used for pattern matching  
\% used to define any number of missing characters before and after a pattern  
\_ wildcard used to substitute for a single character or space  
example:  
Return True if the country name starts with a 'U' else return false  
SELECT country LIKE 'U%'  
FROM MOVIES;  
Return True if country starts with 'U' and ends with 'A' an has one or more spaces in between  
SELECT country LIKE 'U_A'  
FROM MOVIES;  

### _Functions_  
_LOWER()_, _LCASE()_ Return string in lowercase  
_UPPER()_, _UCASE()_ Return string in uppercase  
_CHAR_LENGTH()_ Return number of characters in a
string  
_LEFT(, n)_ Return the leftmost number of
characters specied by n  
_RIGHT(, n)_ Return the rightmost number of
characters specied by n  
_SUBSTR(, m, n)_
Return the number of characters
specied by n starting from the
m-th position of the string  
_LTRIM()_ Return the string with leading
spaces removed  
_RTRIM()_ Return the string with trailing
spaces removed  
_TRIM()_ Return the string with leading
and trailing spaces removed  
_REPLACE('s', 's1', 's2')_
Replaces all the occurrences of a
 within a string with a new
substring  

### _Concatenation_  
CONCAT ( \<col name\>, \<col name\>,...)  

## Date Manipulation  

### Functions
_CURRENT_DATE()_ Return current date.  
_CURRENT_TIME()_ Return current time.  
_DATE()_ Extract the date part of a date or
datetime expression.  
_DAY()_, _DAYOFMONTH()_ Return the day of the month (1-
31)  
_DAYOFWEEK()_ Return the weekday index of a
date or datetime expression.  
_MONTH()_ Return the month part of a date
or datetime expression.  
_YEAR()_ Return the year part of a date or
datetime expression.  
_DATE_ADD(date, INTERVAL value add unit)_  
Adds a time/date interval to a
date and then returns the date  



# Docker

## Images
+ An executable package that includes everything needed to run an application: code, runtime, libraries, environment variable configuration files
+ can build your own 

can find images at https://www.dockerhub.com  


## Container
+ a runtime instance of a image - is what the image becomes in memory when executed
+ once you have the image pulled you can start the container from the docker engine
+ containers isolate software from its environment.  

## Commands
### _List all images_   
docker image ls  

### _Delete a file_  

rm   

### _Continue command on new line_  

^  

### _Pull an image_  

docker pull \<docker IMAGE\>   

### _Remove image from computer_  

docker image rm -f \<IMAGE ID\>   

### _Run a container_  

docker run -it -d --rm ^  
-p hostPort:containerPort ^  
-v host-dir:container-dir ^  
(actual example  
docker run -it --rm ^  
-p 8890:8888 ^  
-v %cd%:/home/ubuntu/Workspace ^  
nycdsa/linux-toolkits) cd is the current directory  
\<docker IMAGE\>  
bash   

### _Options_  

-it runs a container in interactive processes.  
-d (optional) runs a container in the background in detached mode.  
--rm (optional) removes the container when it exits or when the daemon exits.   
-p hostPort:containerPort (optional) forwards a container᾿s port to the host.   
-v host-dir:container-dir (optional) mounts a host directory into a container directory so files can be shared between them.   
bash (optional) starts the command line of the container in the terminal.Depending how the image was built, you may or may not need it.   

### _Inspect running elements_  

docker ps  

### _Inspect stopped containers_  

docker ps -a  

### _Stop a container_  

exit  

### _Remove a stopped container_  

docker container rm <CONTAINER ID>    

only need to specify enough characters of the container to uniquely identify it  

### _Restart and Log back into container_  

docker restart <container id>   

docker exec -it <container id> bash  


# General Windows Terminal 
## Commands

### _Echo_
echo prints  
### _Change Directory_  
cd  
### _list files and directories_  
dir  

# Linux   

## Commands  
date (gets date)  

ssh (loginto remote server)  

scp (copy from or to a remote machine)  

wget or curl (download a webpage)  


### _Command help_  

command --help  

### _Show entire file system_  

ls <directory to list>   


### _Change directories_  

cd <pathname>  

cd or cd ~ changes to home directory  

cd .. changes to parent folder  


### _Make DIrectory_  

mkdir <directory name>  

mkdir -p ~/examples/multiple/levels/down  

(-p makes multiple nested directories regardless of weather parts of the specified path exist)  


### _Copy file or Folder_  

cp [-r] <source> <destination>  

-r argument means recursively for copying folders  

copy a file into home directory; note the .  
$ cp /etc/magic .  
copy a file into home directory with a new name  
$ cp /etc/hosts nethosts  
$ ls  

### _Move or rename files_  
mv <source> <Destination>

### _Move curser and keyboard shortcuts_
ctrl + a (begginging of line)  
crtl + e (end of line)  
crtl + u (delete line)
Tab (auto-complete) 

### _Remove files_
rm
rmdir or rm -r (removes empty folders)  

### _Quick create file_
touch <filename>

### _I/O redirection_
\> create  
\>\> append

### _Print Files or show contents_
cat <filename>
head <filename>  
less <filename>
Scroll up/down one line at a time with arrow keys.  
Use the space bar to scroll through a screen at a time.  
Use the / key to search for a term: /apple  
Press h to see a quick list of other options.  
Press q to exit less.  

### _Download file_
wget <url>

### _File permissions_ 
ls - l (long format) to show permissions  

The permissions are broken into 4 sections. For example: drwxr-xr--  

1. d: ‘-’ -> file, ‘d’ -> directory, ‘l’ -> link  

2. rwx: permissions for the owner  

3. r-x: permissions for members of the group owning the file  

4. r-x: permissions for other users  

Where ‘r’-> read, ‘w’-> write/delete, ‘x’-> execute, ‘-’-> no permission  



### _Show command line format_
echo $PS1  (PS1-4 exist)

### _change command prompt_  
export PS1 = "\\u \\t:$"   

will change prompt to user name and time followed by \"\:\$\"  


### _Clear screen_
crtl + L  

### _Stop process_
crtl + C   

### _Stop Program_
crtl + D  

### _print working dirrectory_
pwd

### _Relative Path_
./  (is current directory)

### _Parent Directory_
../  

### _home directory_
~  

### _show where a file is_
which <filename>  

### _change permission_
write = 2  

execute = 1  

read = 4  

chmod 744  

owner will be given 7 permission (sum of 4 + 1 + 2) 

group will be given 4 permission (sum of 3 + 1)  

other users will be given permission 4 (sum of 3 +1) 


### _Text Processing_
grep - Search through a given file using a string  

grep <word> <file1> [<file2]>  

wc - Count lines and words  

outputs lines/words/characters  

sort - Sort all lines in file  

  sort -k9 (key)(column 9 (counts from 1))  
  
  sort -k5 -n (sorts 5th column which contains numbers)  
  


# VI  
## commands  

### _Open vi_
vi <filename>  
will open filename or create it if it doesnt exist  

### _insert mode_
I  

a - insert after cursor  

i - insert before cursor  

o - insert a new line below  

O - insert a new line above  

ESC - exit insert mode 


### _command mode_
esc  

x - delete one character  
dw - delete one word  
dd - delete the line  
D - delete the rest of the line  
u - undo the last action  

### _exit VI_
ZZ save file and exit(capital letters)  
:wq (save and quit)  
:w save without exiting  
:q! exit without saving  

# Git
## workflow  
-git init  
or  
-git clone  
write code  
-git add <file>  
<file> is now in staging area  
-git commit  
commit changes and opens editor to add a comment  
write code  
-git add <file>  
-git commit -m "messasge" 
revert changed git checkout <hash> or <head~ID>  
brings us to a detached head state  
-git checkout master (or main)  
reattaches head  
for checkouts that could cause conflicts  
-git revert <hashID>  
resolve conflicts  
-git add .  
-git commit -m "message"  
prepare to push to github  
change wd out of project repo  
create repo on github  
copy repo URL from 'code' button  
-git clone <git hub url>  
-git push  




## Commands  
###  _git init_ 
convert an existing, unversioned project to a Git repository or initialize a new empty repository.  

###  _git clone_  
copies an existing Git repository  

###  _git status_ 
show status of all tracked and untracked files 

###  _git log_  
full status of git commands
--stat includes a brief summary  
--oneline (condense each commit to a single line)    
--graph --decorate To include branches and the ref names:  

###  _git reflog_ 
get complete history in repo  

###  _git add <file> _ 

starts tracking <file>  
git add . (adds all files)  

###  _git commit_ 
commit changes  
-m "message"  
-a  does the add as well as commit  

###  _git show_ 
see changes that were made to tracked files  
git show <commit hash> to see textual differences   
git show HEAD reference to the currently checked-out commit/branch.  
git show HEAD~2 shows the second commit message  

###  _git checkout -- <file> _  
discard changes in working directory  
git checkout HEAD~2  

###  _git diff_  
To see what you’ve changed but not yet staged 

###  _git restore --staged <file>  _ 
to unstage <file>  

###  _git reset <hash ID>_ 
permanetly reverts code to hash ID

###  _Config _  
git config --global user.name "name"  
git config --global user.email email@email  



# Resources
Books for free:  Data Science at the Command line  
https://datascienceatthecommandline.com/  
Practical Statistics for Data Science
https://largepdf.net/practical-statistics-for-data-scientists-50-essential-concepts-using-r-and-python/  
Bayesian Methods for hackers
https://freecomputerbooks.com/Bayesian-Methods-for-Hackers.html




